Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-23 ms.tex
gone mad.

\begin{abstract}
  We calculate steady-state, one-dimensional hydrodynamic profiles of
  hot gas in slowly accreting (``quiescent") galactic nuclei for a
  range of central black hole masses $M_{\bullet}$, parameterized gas
  heating rates, and observationally-motivated stellar density
  profiles.  Mass is supplied to the circumnuclear medium by stellar
  winds, while energy is injected primarily by stellar winds,
  supernovae, and black hole feedback.  Analytic estimates are derived
  for the stagnation radius (where the radial velocity of the gas
  passes through zero) and the black hole accretion rate, $\dot{M}$,
  as a function of $M_{\bullet}$ and the gas heating efficiency, the
  latter being related to the star-formation history.  We assess the
  conditions under which radiative instabilities develop, both in the
  case of a single burst of star formation and for the average star
  formation history predicted by cosmological simulations.  By
  combining a sample of measured nuclear X-ray luminosities, $L_{X}$,
  of nearby quiescent galactic nuclei with our results for
  $\dot{M}(M_{\bullet})$ we show that most nuclei are consistent with
  accreting in a steady-state, thermally-stable manner for radiative
  efficiencies predicted by ADAF models.  However, thermally-stable
  accretion cannot explain the short average growth times of low mass
  black holes in the local universe, which must instead result from
  gas being fed in from large radii, due either to galaxy mergers or
  thermal instabilities acting on larger, galactic scales.  Our
  results have implications for attempts to constrain the occupation
  fraction of SMBHs in low mass galaxies using the mean $L_{\rm
    X}-M_{\bullet}$ correlation, as well as the predicted diversity of
  the circumnuclear densities encountered by relativistic outflows
  from tidal disruption events.
\end{abstract}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{mdot_sfr.pdf}
\caption{\label{fig:bh_growth} {\it Top panel:} Black hole accretion
  rate (excluding small-scale outflows) $\dot{M}$ in Eddington units
  as a function of black hole mass $M_{\bullet}$, shown separately for
  core ($\Gamma=0.1$, dot-dashed) and cusp galaxies ($\Gamma=0.8$,
  solid).  Black lines show the accretion rate calculated from
  equation (\ref{eq:eddr_analytic}) using our estimate of the gas
  heating rate provided at $z = 0$ by the average star formation
  histories for each galaxy mass (Fig.~\ref{fig:vwSources}).  Teal
  lines show the maximum accretion set by Ia supernovae blow-out
  (\ref{eq:eddr_Ia}).  Red lines show the accretion rate obtained if
  Compton heating alone was acting (eq.~[\ref{eq:MdotC}]).  Yellow
  lines show the maximum accretion rate for a thermally stable flow
  near the stagnation radius (eq.~[\ref{eq:Mdotmax}]).  We also show
  Eddington ratios inferred for SgrA* from \citet{Quataert:2004a}
  (gray circles), for NGC3115 from \citet{ShcherbakovWong+:2014a}
  (gray diamonds), and for the Swift J1644+57 TDE from
  \citet{BergerZauderer+:2012a} (gray triangles). Note that the actual
  black hole accretion rate will be a factor of the Shakura-Sunyaev
  $\alpha$ smaller than the $\Mdot$s here due to small scale
  outflows. Thus the plot is really showing the gas inflow rate on
  large scales (e.g. near the stagnation radius) {\it Bottom panel:} Growth
  times $\Mbh/\Mdot$ in units of the Hubble time $\th$ for each of the
  accretion rates shown in the top panel.}
\end{figure}

@@

If the galaxies for which $L_{\rm X}$ is measured at low $M_{\bullet}$
in Fig \ref{fig:miller} are indeed in a state of thermally-unstable
outburst, then a subset of the galaxies with X-ray upper limits
may contribute to a separate population of thermally stable, steadily
accreting nuclei.  Such a population could include SgrA*, which has a
much lower X-ray luminosity for its black hole mass than predicted by
the \citet{Miller+15} trend.  A potentially bimodal population of
steady and cyclic galactic nuclei calls into question the physical
justification of using simple extrapolations of power-law fits to the
$L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to
constrain the occupation fraction of SMBHs in the nuclei of low mass
galaxies.

@@

\citet{BergerZauderer+:2012a} infer a flattening of the gas density
profile of the host of J1644+57 on radial scales $r \gtrsim 0.4$ pc
(their Fig.~6) which looks qualitatively similar to the shape density
profile we predict for core galaxies (Fig.~\ref{fig:profiles}, solid
line).  To obtain such a flattening on the inferred radial scales
would require a black hole of $\Mbh\sim 10^7 \Msun$.  However, this is
inconsistent with the break in the gas density seen at $\sim 0.6$ pc
by \citet{BergerZauderer+:2012a}, which should correspond to the break
radius of the {\it stellar density} profile, thus implying an
unrealistically small SMBH mass of $\Mbh\simeq 100\Msun$ according to
the mean $r_b-M_{\bullet}$ relationship
\citep{LauerFaber+:2007a}. {\bf Other constraints on gas density:
  J1644 gone mad}

@@

  \item Including the effects of heat transport by electron conduction
    results in at most order unity changes to the key properties of
    the flow (e.g. stagnation radius, mass accretion rate, and thermal
    stability) for causal values of the conduction saturation
    parameter $\phi < 0.1$ (eq.~[\ref{eq:conduction}]).  Angular
    momentum is also unlikely to invalidate our spherically symmetric
    model, given the inferred stellar rotation rates near the
    stagnation radius ($\S\ref{sec:rotation}$).

@@

  % \item Our calculation of $\Mdot$ agrees well with the one
  %   calculated from the Bondi formula. We find that they agree to
  %   within a factor of order unity. This is because in our model the
  %   accretion rate is determined by the flow properties at the
  %   stagnation radius, $\rs$, which is within a factor of a few of
  %   the
  %   Bondi radius $\rb$. Thus, use of the Bondi formula is a
  %   good approximation.

>>>>

\begin{abstract}
  We calculate steady-state, one-dimensional hydrodynamic profiles of
  hot gas in slowly accreting (``quiescent") galactic nuclei for a
  range of central black hole masses $M_{\bullet}$, parameterized gas
  heating rates, and observationally-motivated stellar density
  profiles.  Mass is supplied to the circumnuclear medium by stellar
  winds, while energy is injected primarily by stellar winds,
  supernovae, and black hole feedback.  Analytic estimates are derived
  for the stagnation radius (where the radial velocity of the gas
  passes through zero) and the black hole accretion rate, $\dot{M}$,
  as a function of $M_{\bullet}$ and the gas heating efficiency, the
  latter being related to the star-formation history.  We assess the
  conditions under which radiative instabilities develop, both in the
  case of a single burst of star formation and for the average star
  formation history predicted by cosmological simulations.  By
  combining a sample of measured nuclear X-ray luminosities, $L_{X}$,
  of nearby quiescent galactic nuclei with our results for
  $\dot{M}(M_{\bullet})$ we show that most nuclei are consistent with
  accreting in a steady-state, thermally-stable manner for radiative
  efficiencies predicted by ADAF models.  However, thermally-stable
  accretion cannot explain the short average growth times of low mass
  black holes in the local universe, which must instead result from
  gas being fed in from large radii, due either to gas inflows or
  thermal instabilities acting on larger, galactic scales.  Our
  results have implications for attempts to constrain the occupation
  fraction of SMBHs in low mass galaxies using the mean $L_{\rm
    X}-M_{\bullet}$ correlation, as well as the predicted diversity of
  the circumnuclear densities encountered by relativistic outflows
  from tidal disruption events.
\end{abstract}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{mdot_sfr.pdf}
\caption{\label{fig:bh_growth} {\it Top panel:} Gas inflow rate
  (excluding small-scale outflows) $\dot{M}$ in Eddington units as a
  function of black hole mass $M_{\bullet}$, shown separately for core
  ($\Gamma=0.1$, dot-dashed) and cusp galaxies ($\Gamma=0.8$, solid).
  Black lines show the accretion rate calculated from equation
  (\ref{eq:eddr_analytic}) using our estimate of the gas heating rate
  provided at $z = 0$ by the average star formation histories for each
  galaxy mass (Fig.~\ref{fig:vwSources}).  Teal lines show the maximum
  accretion set by Ia supernovae blow-out (\ref{eq:eddr_Ia}).  Red
  lines show the accretion rate obtained if Compton heating alone was
  acting (eq.~[\ref{eq:MdotC}]).  Yellow lines show the maximum
  accretion rate for a thermally stable flow near the stagnation
  radius (eq.~[\ref{eq:Mdotmax}]).  We also show Eddington ratios
  inferred for SgrA* from \citet{Quataert:2004a} (gray circles), for
  NGC3115 from \citet{ShcherbakovWong+:2014a} (gray diamonds), and for
  the Swift J1644+57 TDE from \citet{BergerZauderer+:2012a} (gray
  triangles). Note that the actual black hole accretion rate could one
  to several orders of magnitude smaller due to the presence of small
  scale outflows. Thus the plot is really showing
  the gas inflow rate on large scales (e.g. near the stagnation
  radius) {\it Bottom panel:} Growth times $\Mbh/\Mdot$ in units of
  the Hubble time $\th$ for each of the accretion rates shown in the
  top panel.}
\end{figure}

@@

If the galaxies for which $L_{\rm X}$ is measured at low $M_{\bullet}$
in Fig \ref{fig:miller} are indeed in a state of thermally-unstable
outburst, then a subset of the galaxies with X-ray upper limits
may contribute to a separate population of thermally stable, steadily
accreting nuclei.  Such a population could include SgrA*, which has a
much lower X-ray luminosity for its black hole mass than predicted by
the \citet{Miller+15} trend.  A potentially bimodal population of
steady and outbursting galactic nuclei calls into question the physical
justification of using simple extrapolations of power-law fits to the
$L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to
constrain the occupation fraction of SMBHs in the nuclei of low mass
galaxies.

@@

\citet{BergerZauderer+:2012a} infer a flattening of the gas density
profile of the host of J1644+57 on radial scales $r \gtrsim 0.4$ pc
(their Fig.~6) which looks qualitatively similar to the shape density
profile we predict for core galaxies (Fig.~\ref{fig:profiles}, solid
line).  To obtain such a flattening on the inferred radial scales
would require a black hole of $\Mbh\sim 10^7 \Msun$.  However, this is
inconsistent with the break in the gas density seen at $\sim 0.6$ pc
by \citet{BergerZauderer+:2012a}, which should correspond to the break
radius of the {\it stellar density} profile, thus implying an
unrealistically small SMBH mass of $\Mbh\simeq 100\Msun$ according to
the mean $r_b-M_{\bullet}$ relationship \citep{LauerFaber+:2007a}. We
also note that other constraints (the observed jet power, the jet
shut-off (if linked with the accretion becoming sub-Eddington), and an
observed QPO in the x-ray light curve) favor a smaller black hole
($10^5-10^6 \Msun$) \citep{Tchekhovskoy+2014}.

@@

  \item Including the effects of heat transport by electron conduction
    results in at most order unity changes to the key properties of
    the flow (e.g. stagnation radius, mass accretion rate, and thermal
    stability) for causal values of the conduction saturation
    parameter $\phi < 0.1$ (eq.~[\ref{eq:conduction}]).  Angular
    momentum is also unlikely to invalidate our spherically symmetric
    model on large scales, given the inferred stellar rotation rates
    near the stagnation radius ($\S\ref{sec:rotation}$). However, on
    small scales gas would encounter a centrifugal barrier which would
    cause much of the gas flowing inwards on large scales to be driven
    outwards in an outflow (as in \citealt{Li+2013}).

@@
